---
title: "Hands-on Exercise 2: Choropleth Mapping with R"
format: 
  html:
    code-line-numbers: true
    number-sections: true
    highlight-style: github
execute:
  warning: false
date: "`r Sys.Date()`"
---

# Getting Started

In this exercise, I hope to accomplish the following:

-   under choropleth mapping involves the symbolisation of enumeration units (e.g., countries, provinces, using area patterns or graduated colors), and
-   how to plot choropleth maps using the `tmap` package.

# Packages Used

The following packages will be used in this exercise:

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/) for plotting choropleth maps,
-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,
-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,
-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and
-   [**sf**](https://cran.r-project.org/web/packages/sf/) for handling geospatial data.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Data

## Datasets used

-   **Geospatial:** [Master Plan 2014 Subzone Boundary (Web)](https://data.gov.sg/dataset/master-plan-2014-subzone-boundary-web)

-   **Aspatial:** [Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) in csv format

## Importing Geospatial Data

```{r}
mpsz <- st_read(dsn="data/geospatial",
                layer="MP14_SUBZONE_WEB_PL")
mpsz
```

## Importing Aspatial Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## Data Preparation

A data table with year 2020 values is needed for the thematic map. In this table, variables PA - Planning Area, SZ - Subzone, and AG - Age Group (ECONOMY ACTIVE, AGED, TOTAL, and DEPENDENCY) will be required.

-   **YOUNG:** age group 0 to 4 until age group 20 to 24,
-   **ECONOMY ACTIVE:** age group 25-29 until age group 60-64,
-   **AGED:** age group 65 and above,
-   **TOTAL:** all age group, and
-   **DEPENDENCY:** the ratio between young and aged against economy active group

### Data Wrangling

To wrangle the data, these functions will be used:

-   `pivot_wider()` of **tidyr** package, and
-   `mutate()`, `filter()`, `group_by()` and `select()` of **dplyr** package

```{r}
popdata2020 <- popdata %>%
                filter(`Time` == 2020) %>%
                group_by(PA, SZ, AG) %>% 
                # creates a new data frame with one (or more) rows for each combination of grouping variables
                # get total population for each group
                summarise(`POP` = sum(`Pop`)) %>%
                # ungroup() is used after the group() command after performing calculations to create summaries for each grouping
                ungroup() %>%
                # increase number of columns and decrease the number of rows
                pivot_wider(names_from = AG, # name of output column
                             values_from = POP) %>% # column(s) to obtain cell values from
                #  mutate() adds new variables and preserves existing ones
                # rowSums() returns the sums of each row 
                mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>% # get the columns with the age ranges
                mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
                mutate(`AGED` = rowSums(.[16:21])) %>%
                mutate(`TOTAL` = rowSums(.[3:21])) %>%
                mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)/`ECONOMY ACTIVE`) %>%
                select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)


popdata2020
```

### Joining Aspatial Data with Geospatial Data

Before performing georelational join, PA and SZ fields needs to be converted into uppercase. This is due to the values in PA and SZ fields having both upper- and lowercases. Whereas SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
                mutate_at(.vars = vars(PA, SZ),
                          .funs = funs(toupper)) %>%
                filter(`ECONOMY ACTIVE` > 0)
popdata2020
```

Next, `left_join()` of **dplyr** is used to join the geographical data and attribute table using planning subzone name (i.e., *SUBZONE_N* and *SZ* use the common identifier).

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
mpsz_pop2020
```

::: callout-tip
In the use of `left_join()`, mpsz simple feature dataframe is used as the left data table to ensure that the output will be a simple features dataframe.
:::

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

::: callout-caution
Remember to create a "rds" folder inside the "data" folder.
:::

## Choropleth Mapping Geospatial Data using *tmap*

Two approaches can be used to prepare thematic map using *tmap*,:

-   Plotting a thematic map quickly by using `qtm()`.
-   Plotting highly customisable thematic map by using tmap elements.

### Plotting a choropleth map quickly by using `qtm()`

The easiest and quickest to draw a choropleth map using ***tmap*** is using `qtm()`. It is concise and provides a good default visualisation in many cases.

The code chunk below will draw a cartographic standard choropleth map as shown below.

```{r}
tmap_mode("plot") # set to plotting
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

::: callout-note
-   `tmap_mode()` with "plot" option is used to produce a static map. For interactive mode, "view" option should be used.
-   *fill* argument is used to map the attribute (i.e. DEPENDENCY)
:::

### Creating a choropleth map by using *tmap*'s elements

Despite drawing choropleth map quickly and easily, `qtm()` makes aesthetics of individual layers harder to control. To draw a high quality cartographic choropleth map as shown in the figure below, **tmap**'s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) + # transparency
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

#### Drawing a base map

The basic building block of **tmap** is `tm_shape()` followed by one or more layer elemments such as `tm_fill()` and `tm_polygons()`.

In the code chunk below, `tm_shape()` is used to define the input data (i.e *mpsz_pop2020*) and `tm_polygons()` is used to draw the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### Drawing a choropleth map using *tm_polygons()*

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, the target variable (e.g., *Dependency*) needs to be assigned to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

::: callout-note
-   The default interval binning used to draw the choropleth map is called "pretty".
-   The default colour scheme used is `YlOrRd` of ColorBrewer.
-   By default, missing value will be shaded in grey.
:::

#### Drawing a choropleth map using `tm_fill()` and `tm_border()`

`tm_polygons()` is a wraper of `tm_fill()` and `tm_border()`. `tm_fill()` shades the polygons using the default colour scheme and `tm_borders()` adds the borders of the shapefile onto the choropleth map.

The code chunk below draws a choropleth map by using *tm_fill()* alone.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") # no borders
```

::: callout-note
Notice that the planning subzones are shared according to the respective dependecy values
:::

To add the boundary of the planning subzones, `tm_borders()` is used as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1, alpha = 1)
```

With the addition of `tm_borders()`, light-gray border lines have been added on the choropleth map.

There are four arguments for `tm_borders()`:

-   *alpha* = transparency. It ranges between 0 (totally transparent) and 1 (not transparent). By default, the alpha value is 1.
-   *col* = border colour,
-   *lwd* = border line width. The default is 1, and
-   *lty* = border line type. The default is "solid".

### Data classification methods of **tmap**

Most choropleth maps employ some methods of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

**tmap** provides ten data classification methods, namely: *pretty* (default), *equal*, *jenks*, *fixed*, *sd*, *quantile*, *kmeans*, *hclust*, *bclust*, and *fisher*.

To define a data classification method, the *style* argument of `tm_fill()` or `tm_polygons()` will be used.

#### Plotting choropleth maps with built-in classification methods

##### Classification Methods

The following code chunks will showcase the different data classifications methods using 5 classes.

###### **Jenks** **Natural Breaks - *Jenks***

The code chunk below shows a quantile data classification that used 5 classes:

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

###### *Equal*

In the code chunk below, *equal* data classification method is used.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

::: callout-note
Notice that the distribution of quantile data classification method are *more evenly distributed* then equal data classification method.
:::

###### *Fixed*

*fixed* enables user to set their own desired ranges.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fixed",
          breaks = c(0, 2, 6, 13, 19)) + 
  tm_borders(alpha = 0.5)
```

###### Standard Deviations - *sd*

*sd* calculates standard deviations of a given variable, and use these values as the break widths.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)
```

###### *Quantile*

This style creates breaks with an equal number of features (polygons).

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

###### *KMeans*

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

###### **Hierarchical clustering - *hclust***

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "hclust") +
  tm_borders(alpha = 0.5)
```

###### **Bagged Clustering - *bclust***

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "bclust") +
  tm_borders(alpha = 0.5)
```

###### Fisher Algorithm - *fisher*

"fisher" creates groups with maximalized homogeneity.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fisher") +
  tm_borders(alpha = 0.5)
```

##### Different Number of Classes

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 20,
          style = "pretty") +
  tm_borders(alpha = 0.5)
```

#### Plotting choropleth map with custom breaks

For all the built-in styles, the category breaks are **computed internally**. To override these defaults, the breakpoints can be set explicitly by means of the *breaks* argument to the `tm_fill()`. It is important to note that, in **tmap** the breaks include a minimum and maximum. As a result, in order to end up with n categories, n+1 elements must be specified in the *breaks* option (the values must be in increasing order).

::: callout-tip
It is always a good practice to get some descriptive statistics on the variable before setting the break points.
:::

Code chunk below will be used to compute and display the descriptive statistics of **DEPENDENCY** field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the results above, break points will be set to 0.60, 0.70, 0.80, and 0.90. As mentioned, a minimum and a maximum need to be stated, which are set as 0 and 1 here. The *breaks* vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

Now, the choropleth map is coded as shown below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.6, 0.7, 0.8, 0.9, 1)) +
  tm_borders(alpha = 0.5)
```

### Colour Schemes

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the **RColorBrewer** package.

##### Using ColourBrewer palette

To change the colour, the preferred colour needs to be assigned to the *palette* argument of `tm_fill()` as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

Here, the palette is *green*.

To reverse the colour shading, add a "-" prefix to *green*.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

### Map Layouts

Map layout refers to the combination of all map elements into a cohensive map. Map elements include among others the objects to be mapped, the title, the scale bar, the compass, margins and aspects ratios.

Colour settings and data classification methods covered in the previous section relate to the palette and break-points are used to affect how the map looks.

#### Map Legends

In **tmap**, several *legend* options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "jenks",
          palette = "Blues",
          legend.hist = TRUE, # logic that determines whether a histogram is shown
          legend.is.portrait = TRUE, # legend is portrait or landscape
          legend.hist.z = 0.1) + # index value that determines the position of the histogram legend element
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45,
            legend.width = 0.35,
            legend.outside = FALSE, # whether the legend is plot outside of the map
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

#### Map Style

**tmap** allows a wide variety of layout settings to be changed. They can be called by using *tmap_style()*.

The code chunk below shows the *classic* style is used.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### Cartographic Furniture

Beside map style, **tmap** also also provides arguments to draw other map furniture such as compass, scale bar and grid lines.

In the code chunk below, `tm_compass()`, `tm_scale_bar()` and `tm_grid()` are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          title = "No. of Persons") + # title for ranges
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) + # 	line width of the grid lines
   tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
  
```

### Drawing Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arrange side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in three ways:

-   by assigning multiple values to at least one of the aesthetic arguments,
-   by defining a group-by variable in *tm_facets()*, and
-   by creating multiple stand-alone maps with *tmap_arrange()*.

#### By assigning multiple values to at least one of the aesthetic arguments

In this example, small multiple choropleth maps are created by defining ***ncols*** in **tm_fill()**

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(c("DEPENDENCY", "AGED"),
              style = c("equal", "quantile"),
              palette = list("Blues", "Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in `tm_facets()`

In this example, multiple small choropleth maps are created by using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by = "REGION_N", # data variable name by which the data is split
            # If "by" is specified, should each map has its own coordinate ranges? By default TRUE, unless facets are shown in as different layers
            free.coords = TRUE, 
            # If "by" is specified, should non-selected spatial units be dropped? If FALSE, they are plotted where mapped aesthetics are regarded as missing values.
            drop.shapes = TRUE) + 
  tm_layout(legend.show = FALSE, 
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with `tmap_arrange()`

In this example, multiple small choropleth maps are created by creating multiple stand-alone maps with `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap,
             asp = 1, # aspect ratio
             ncol = 2) # number of columns
```

### Mappping Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth map, you can also use selection funtion to map spatial objects meeting the selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
